---
title: "regression, final project"
author: "Laura Lynch"
date: "11/26/2019"
output: html_document
---

```{r}
# set up
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(viridis)
library(leaflet)
library(dplyr)
library(corrplot)
library(modelr)
library(mgcv)
library(purrr)
library(stringr)


knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 10, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r}
# original cleaning code

drug_death=
  read_csv("./data/Accidental_Drug_Related_Deaths_2012-2018.csv") %>%
  janitor::clean_names() %>%
  separate(date, into = c("date", "time"), sep = 11) %>%
  select(-time) %>%
  separate(date, into = c("month", "day", "year"), sep = " , ", convert = TRUE) %>%
  drop_na(age, sex, race, residence_county, residence_state) %>%
  mutate(death_city_longlat = sub("\\).*", "", sub(".*\\(", "", death_city_geo))) %>%
  # Laura changed date to month
  separate(month, into = c("month", "day", "year"), sep = "/", convert = TRUE, remove = FALSE) %>%
  # Laura deleted:      date = as.Date(date, format = '%m/%d/%Y'), 
  mutate(
     death_city_longlat = sub("\\).*", "", sub(".*\\(", "", death_city_geo))  ) %>% 
  separate(col = death_city_longlat, into = c("death_city_lat", "death_city_long"), sep = "," ,remove = FALSE) %>%
  mutate(
      injury_city_longlat = sub("\\).*", "", sub(".*\\(", "", injury_city_geo))  ) %>% 
      separate(col = injury_city_longlat, into = c("injury_city_lat", "injury_city_long"), sep = "," ,remove = FALSE) %>%
  mutate(
      residence_city_longlat = sub("\\).*", "", sub(".*\\(", "", residence_city_geo))  ) %>% 
      separate(col = residence_city_longlat, into = c("residence_city_lat", "residence_city_long"), sep = "," ,remove = FALSE) %>%
  ##dropping pre-split variables for latitude and longitude, dropping date_type because it is uninformative for this analysis 
  select(-residence_city_longlat, -death_city_longlat, -injury_city_longlat, -residence_city_geo, -injury_city_geo, -death_city_geo, -mannerof_death, -date_type) %>%
  mutate(race = recode(race,
                       "Asian Indian" = "Asian",
                       "Asian, Other" = "Asian",
                       "Chinese" = "Asian",
                       "Hawaiian" = "Other"),
         race = na_if(race, "Unknown")) 
  
drug_death
```

```{r}
CT = c("LITCHFIELD", "HARTFORD", "TOLLAND", "NEW LONDON", "MIDDLESEX", "NEW HAVEN", "WINDHAM", "FAIRFIELD")

drug_death_bin =
  drug_death %>%
  # binarize outcome: death in hospital YES or NO
  mutate(death_location = ifelse(location == "Hospital", c(1), c(0))) %>%
  # narrow data set to relevent regression variables
  select(death_location, age, sex, race, residence_county) %>%
  # identify out of state counties
  mutate(new_county = ifelse(residence_county %in% CT,
                             str_to_lower(residence_county, locale = "en"),
                             "out of state")) %>%
  select(-residence_county) %>%
  drop_na(death_location, race) %>%
  # create factor variables for regression
  mutate(death_location = factor(death_location, levels = c(1, 0), labels = c("In hospital", "Outside hospital")),
        sex = factor(sex, labels = c("Male", "Female", "Unknown")),
        race = factor(race, labels = c("White", "Black", "Hispanic, White", "Asian", "Other", "Hispanic, Black")),
        new_county = factor(new_county, labels = c("litchfield", "hartford", "tolland", "new london", "middlesex", "new haven", "windham", "fairfield", "out of state"))
          ) 

drug_death_bin
```

```{r}
# specify predictor variable types
drug_death_bin %>%  
  mutate(
    age = as.numeric(age),
    sex = fct_infreq(sex),
    race = fct_infreq(race),
    new_county = fct_infreq(new_county)
    )

# create model
model=
  glm(death_location ~ age + sex + race + new_county, data = drug_death_bin, family = "binomial") 


model %>%
  broom::glance()%>% 
  knitr::kable(digits = 3)

model %>%
  broom::tidy()%>% 
  knitr::kable(digits = 3)

summary(model)
# significant p = age, sexFemale, counties: Hartford, New London, Out of State

model %>% 
  broom::tidy() %>% 
  mutate(OR = exp(estimate)) %>%
  select(term, log_OR = estimate, OR, p.value) %>% 
  knitr::kable(digits = 3)
```

# Analysis

In order to understand the likelihood of a drug related death occuring in the hospital versus not in the hospital, we binarized the variable `death_location` as the outcome (1 = In the hospital, 0 = Not in the hospital). For this logistic regression analysis we used a generlized logistic model with the predictors age, sex, race and county of residence. `Age` was treated as a continuous variable. `Sex` was categorical (Male, Female, and Unknown) with Male as the reference category based on frequency. `Race` was categorical with White, Black, Hispanic White, Asian, Other, Hispanic Black; White was the reference category based on frequency and NAs were dropped pre-analysis (n= 16). Finally the 8 `Counties` of CT were converted to factor variables with all other counties being converted to "out of state" with Litchfield Country as the reference variable based on frequency, NAs were dropped (n= 7).

The significant variables in the model (p-value < 0.05) were  age, Female, Hartford and New London Counties and the  "Out of State" category.
TERM          OR      P-VALUE Odds of dying in the hospital vs. outside the hospital
age           1.006   0.049   (for every 1-year increase in age)
female        1.207   0.028   (vs. Male)
Hartford      1.497   0.000   (vs. Litchfield)
New London    1.769   0.007   "
Out of State  1.741   0.015   "
